Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-11 appendix.tex
I was more explicit in the theorem about cases that were exceptions

%das: TODO double check restrictions on \hat{x} and \hat{y} are reasonable
Since $x^*,y^*,\hat{x},$ and $\hat{y}$ all fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)

@@

\beta > \gamma.

@@

@@

Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula.

>>>>

%das: added as assumption to theorem
Since $x^*,y^*,\hat{x},$ and $\hat{y}$ all fall between zero and one, we make use of the inequalities

@@

\beta > \gamma

@@

using the assumptions that $x^*=y^*=0$, $\hat{x}=\hat{y}=0$, and $\hat{x}=\hat{y}=1$.

@@

Using the assumption that $\hat{x}\neq 0$ and $\hat{y}\neq 0$, this implies $\ell'$ has two zeros according to the quadratic formula.

### 2018-05-11 joint_inf.tex
clarify approximation/bias
I was more explicit in the theorem about cases that were exceptions

%Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
%We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
The estimate $\hat{w}$ in \eqref{eq:general-bl-opt} is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
We fix bounds on these estimates and show that, in optimizing \eqref{eq:general-bl-opt}, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define
\[
\gamma(x, y) = 4xy,
\]
\[
\beta(x, y) = 1+x^2+y^2+x^2y^2,
\]
and bounds
\[
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
\]
\[
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
\]
and
\[
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}).
\]
The maximum likelihood value $\hat{w}=1$ when
\[
-\gamma_{U}^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
\]
\end{restatable}
We use this theorem to show incorrect branch parameter estimates as follows.
If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$; this corresponds to the restricted case, analyzed in the Appendix (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
\[
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
\]
and
\[
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
\]
Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2017-jo} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference converges on the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2017-jo}, namely their Figures~2 and 3.

>>>>

%Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
%We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
The estimate $\hat{w}$ in \eqref{eq:general-bl-opt} is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
We fix bounds on these estimates and show that, in optimizing \eqref{eq:general-bl-opt}, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define
\[
\gamma(x, y) = 4xy,
\]
\[
\beta(x, y) = 1+x^2+y^2+x^2y^2,
\]
and bounds
\[
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
\]
\[
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
\]
and
\[
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}).
\]
Assume $\hat{x}\neq 0,\hat{y}\neq 0,$ $\hat{x}$ and $\hat{y}$ are not both simultaneously one, and $x^*$ and $y^*$ are not both simultaneously zero.
Then, the maximum likelihood value $\hat{w}=1$ when
\[
-\gamma_{U}^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
\]
\end{restatable}
We use this theorem to show incorrect branch parameter estimates as follows.
If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$; this corresponds to the restricted case, analyzed in the Appendix (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
\[
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
\]
and
\[
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
\]
Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2017-jo} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference converges on the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2017-jo}, namely their Figures~2 and 3.
\citet{Sagulenko2017-jo} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:bl-general-bias} shows even when branches are short there is a consistent bias.
